Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-17 paper.tex
Minor edits

\paragraph{Lazy Insertion.}
Adding a hypothesis to a rule state is fairly cheap, but not free: it involves one unification for each premise that the hypothesis likely matches (according to a discrimination tree index), followed by several operations on the variable indices.
We therefore delay insertions until the index has selected at least one hypothesis for each slot.
This helps, for example, with the defining rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
When this rule is used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but with lazy insertion we only process $a$ if there is, in fact, an involutive function is the context.

@@

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of hypotheses},
        ylabel={Time in ms},
        xmin=1, xmax=16,
        ymin=2^3, ymax=2^13,
        xtick={1,2,4,8,16},
        ytick={2^3,2^5,2^7,2^9,2^11,2^13},
        xmode=log,
        ymode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Transitivity benchmark}
    \label{fig:trans}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of Rules},
        %ylabel={Time in ms},
        xmin=1, xmax=32,
        ymin=2^3, ymax=1700,
        ymode=log,xmode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]

        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 10.399958) (2, 16.426541) (4, 43.461403) (8, 205.595541) (16, 1238.090889) (32, 8927.857819)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 11.231875) (2, 12.879500) (4, 16.851930) (8, 24.666485) (16, 42.054055) (32, 76.369778)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 19.405222) (2, 33.487472) (4, 80.107597) (8, 284.954791) (16, 1536.089624) (32, 10721.729861)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.892069) (2, 40.031930) (4, 69.446902) (8, 130.640986) (16, 260.653917) (32, 481.827902)
          };
      \end{axis}
    \end{tikzpicture}
    % graph \includegraphics[width=.4\linewidth]{image1}
    \caption{Independence benchmark}
    \label{fig:indep}
  \end{subfigure}
  \caption{The above results depict the behavior of both algorithms in a context where there is a lot of work that can be reused (Transitivity) and no work to be reused (Independence). The lines with square entry points represent problems where unification is easy. The lines with circle entry points are the same tests with hard unification problems.}
  \label{fig:benchmark}
  \end{figure}

@@

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        title={Depth Test},
        xlabel={Depth},
        ylabel={Time in ms},
        xmin=1, xmax=6,
        ymin=0, ymax=2^10,
        xtick={1,2,3,4,5,6},
        ytick={4,16,64,256,1024},
        ymode=log,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 32.580083) (2, 33.212805) (3, 40.239833) (4, 47.670277) (5, 54.647458) (6, 61.044958) 
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 127.925541) (2, 140.497597) (3, 123.719264) (4, 130.819361) (5, 121.047667) (6, 123.789500)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 108.120444) (2, 107.352694) (3, 116.182833) (4, 125.995639) (5, 137.815527) (6, 142.310597) 
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 604.595791) (2, 624.541194) (3, 608.671111) (4, 614.285014) (5, 588.636819) (6, 595.276806) 
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Depth benchmark no-precompilation}
    \label{fig:erase-noprecomp}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        title={Depth Test},
        xlabel={Depth},
        %ylabel={Time in ms},
        xmin=1, xmax=6,
        ymin=0, ymax=170,
        xtick={1,2,3,4,5,6},
        ytick={2^1,2^3,2^5,2^7},
        ymode=log,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 12.820514) (2, 12.645666) (3, 17.582138) (4, 22.730528) (5, 27.280097) (6, 32.578722) 
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 13.406416) (2, 13.747139) (3, 12.503764) (4, 12.950486) (5, 12.282083) (6, 12.151930) 
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 83.032819) (2, 79.396889) (3, 85.939180) (4, 92.495542) (5, 96.461291) (6, 105.517458)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.681583) (2, 25.322444) (3, 27.672833) (4, 26.999055) (5, 25.690319) (6, 25.257291)
          };
      \end{axis}
    \end{tikzpicture}
    % graph \includegraphics[width=.4\linewidth]{image1}
    \caption{Depth benchmark with precompilation}
    \label{fig:erase-precomp}
  \end{subfigure}
  \caption{The above results depict the behavior of both algorithms in depending on the \text{depth} of potential partial matches. The lines with square entry points represent problems where unification is easy. The lines with circle entry points are the same tests with hard unification problems.}
  \label{fig:erase}
  \end{figure}

>>>>

\paragraph{Lazy Insertion.}
Adding a hypothesis to a rule state is fairly cheap, but not free: it involves one unification for each slot that the hypothesis likely matches (according to a discrimination tree index), followed by several operations on the variable indices.
We therefore delay insertions until the index has selected at least one hypothesis for each slot.
This helps, for example, with the defining rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
When this rule is used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but with lazy insertion we only process $a$ if there is, in fact, an involutive function is the context.

@@

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of hypotheses},
        ylabel={Time in ms},
        xmin=1, xmax=16,
        ymin=2^3, ymax=2^13,
        xtick={1,2,4,8,16},
        ytick={2^3,2^5,2^7,2^9,2^11,2^13},
        xmode=log,
        ymode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Transitivity benchmark}%
    \label{fig:trans}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of Rules},
        %ylabel={Time in ms},
        xmin=1, xmax=32,
        ymin=2^3, ymax=1700,
        ymode=log,xmode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]

        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 10.399958) (2, 16.426541) (4, 43.461403) (8, 205.595541) (16, 1238.090889) (32, 8927.857819)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 11.231875) (2, 12.879500) (4, 16.851930) (8, 24.666485) (16, 42.054055) (32, 76.369778)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 19.405222) (2, 33.487472) (4, 80.107597) (8, 284.954791) (16, 1536.089624) (32, 10721.729861)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.892069) (2, 40.031930) (4, 69.446902) (8, 130.640986) (16, 260.653917) (32, 481.827902)
          };
      \end{axis}
    \end{tikzpicture}
    % graph \includegraphics[width=.4\linewidth]{image1}
    \caption{Independence benchmark}%
    \label{fig:indep}
  \end{subfigure}
  \caption{The above results depict the behavior of both algorithms in a context where there is a lot of work that can be reused (Transitivity) and no work to be reused (Independence). The lines with square entry points represent problems where unification is easy. The lines with circle entry points are the same tests with hard unification problems.}%
  \label{fig:benchmark}
  \end{figure}

@@

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        title={Depth Test},
        xlabel={Depth},
        ylabel={Time in ms},
        xmin=1, xmax=6,
        ymin=0, ymax=2^10,
        xtick={1,2,3,4,5,6},
        ytick={4,16,64,256,1024},
        ymode=log,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 32.580083) (2, 33.212805) (3, 40.239833) (4, 47.670277) (5, 54.647458) (6, 61.044958) 
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 127.925541) (2, 140.497597) (3, 123.719264) (4, 130.819361) (5, 121.047667) (6, 123.789500)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 108.120444) (2, 107.352694) (3, 116.182833) (4, 125.995639) (5, 137.815527) (6, 142.310597) 
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 604.595791) (2, 624.541194) (3, 608.671111) (4, 614.285014) (5, 588.636819) (6, 595.276806) 
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Depth benchmark no-precompilation}%
    \label{fig:erase-noprecomp}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        title={Depth Test},
        xlabel={Depth},
        %ylabel={Time in ms},
        xmin=1, xmax=6,
        ymin=0, ymax=170,
        xtick={1,2,3,4,5,6},
        ytick={2^1,2^3,2^5,2^7},
        ymode=log,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 12.820514) (2, 12.645666) (3, 17.582138) (4, 22.730528) (5, 27.280097) (6, 32.578722) 
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 13.406416) (2, 13.747139) (3, 12.503764) (4, 12.950486) (5, 12.282083) (6, 12.151930) 
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 83.032819) (2, 79.396889) (3, 85.939180) (4, 92.495542) (5, 96.461291) (6, 105.517458)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.681583) (2, 25.322444) (3, 27.672833) (4, 26.999055) (5, 25.690319) (6, 25.257291)
          };
      \end{axis}
    \end{tikzpicture}
    % graph \includegraphics[width=.4\linewidth]{image1}
    \caption{Depth benchmark with precompilation}%
    \label{fig:erase-precomp}
  \end{subfigure}
  \caption{The above results depict the behavior of both algorithms in depending on the \text{depth} of potential partial matches. The lines with square entry points represent problems where unification is easy. The lines with circle entry points are the same tests with hard unification problems.}%
  \label{fig:erase}
  \end{figure}
